---
title: "CSMOM"
author: "Jeronimo Munoz"
date: "2025-07-30"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 

# Cross-Sectional Momentum (CSMOM)

```{r}
# Load the Libraries
library(readxl)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(purrr)
library(broom)

library(caret)
library(randomForest)
library(xgboost)
library(zoo)
library(pROC)
```

## Country 1: Brazil 🇧🇷

### Step 1 – Load and Prepare Brazil Data

```{r}
# importing historical price data for Brazilian stocks 

brazil <- read_excel("Brazil_IBOV_ Data.xlsx", na = c("N/A", "#N/A"))

```

### Step 2 – Format Dates and Columns

```{r}
brazil <- brazil %>%                  
  rename(Date = 1) %>%                # Rename first column to 'Date'
  mutate(Date = as.Date(Date)) %>%    # Convert to Date format
  arrange(Date)                       # Sort data chronologically
```

### Step 3 – Convert Price Columns to Numeric

```{r}
# Convert all price columns to numeric (excluding Date)
brazil_prices <- brazil %>%
  mutate(across(-Date, ~ as.numeric(.)))
```

## Step 4 – Calculate Momentum (t-2 / t-12 - 1)

```{r}

# Calculate momentum: (Price at t-2 / Price at t-12) - 1
momentum_brazil <- brazil_prices %>%
  select(-Date) %>%                                          # Exclude Date column
  mutate(across(everything(), ~ lag(., 2) / lag(., 12) - 1)) # Momentum formula

# Reattach the Date column
momentum_brazil <- bind_cols(Date = brazil$Date, momentum_brazil)
```

### Step 5 – Rank into Deciles

```{r}
# Reshape to long format and assign momentum deciles per date
momentum_deciles_brazil <- momentum_brazil %>%
  pivot_longer(-Date, names_to = "Stock", values_to = "Momentum") %>%
  group_by(Date) %>%
  mutate(
    Decile = ntile(Momentum, 10)  # Rank into 10 groups
  ) %>%
  ungroup()
```

### Step 6 – Calculate 1-Month Forward Returns

```{r}
# 1. Remove Date column and calculate 1-month forward returns
returns_forward_brazil <- brazil %>%
  select(-Date) %>%
  mutate(across(everything(), ~ as.numeric(.))) %>%
  mutate(across(everything(), ~ lead(.) / . - 1))

# 2. Add back the Date column
returns_forward_brazil <- bind_cols(Date = brazil$Date, returns_forward_brazil)
```

### Step 7 – Merge Momentum & Returns

```{r}
# Reshape forward returns to long format (one row per stock-date)
returns_long_brazil <- returns_forward_brazil %>%
  pivot_longer(-Date, names_to = "Stock", values_to = "ForwardReturn")

# Merge forward returns with momentum decile information
final_data_brazil <- momentum_deciles_brazil %>%
  left_join(returns_long_brazil, by = c("Date", "Stock"))
```

### Step 8 – Summary Statistics and Visualisation

```{r}
# Calculate average 1-month forward return per momentum decile (across all dates)
decile_summary_brazil <- final_data_brazil %>%
  group_by(Decile) %>%
  summarise(
    AvgForwardReturn = mean(ForwardReturn, na.rm = TRUE), # Mean return by decile
    Count = n()                                           # Number of observations 
  )
```

```{r}
# Bar chart of average forward return by momentum decile
ggplot(decile_summary_brazil, aes(x = Decile, y = AvgForwardReturn)) +
  geom_col(fill = "steelblue") +
  labs(title = "Average 1-Month Forward Return by Momentum Decile",
       x = "Momentum Decile",
       y = "Avg Forward Return") +
  theme_minimal()
```

### Step 9 – Identify Top and Bottom Decile Stocks (Latest Date)

```{r}
# Get the latest date available in the data
latest_date_brazil <- max(momentum_deciles_brazil$Date, na.rm = TRUE)

# Extract top decile (highest momentum) stocks for that date
top_decile_stocks_brazil <- momentum_deciles_brazil %>%
  filter(Date == latest_date_brazil, Decile == 10) %>%
  select(Stock, Momentum)

# Extract bottom decile (lowest momentum) stocks for that date
bottom_decile_stocks_brazil <- momentum_deciles_brazil %>%
  filter(Date == latest_date_brazil, Decile == 1) %>%
  select(Stock, Momentum)

# Print the stocks in each group
cat("Top Decile (Winners) on", latest_date_brazil, ":\n")
print(top_decile_stocks_brazil)

cat("\nBottom Decile (Losers) on", latest_date_brazil, ":\n")
print(bottom_decile_stocks_brazil)

```

### Step 10 – Long–Short Strategy

```{r}
# Calculate long–short strategy returns: Decile 10 (long) minus Decile 1 (short)
long_short_spread_brazil <- final_data_brazil %>%
  filter(Decile %in% c(1, 10)) %>%    # Keep only top and bottom deciles
  group_by(Date, Decile) %>%
  summarise(Return = mean(ForwardReturn, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Decile, values_from = Return, names_prefix = "Decile") %>%
  mutate(LongShort = Decile10 - Decile1) %>%   # Strategy return
  arrange(Date)
```

```{r}
# Plot cumulative return of long–short strategy over time
long_short_spread_brazil <- long_short_spread_brazil %>%
  mutate(CumulativeReturn = cumsum(replace_na(LongShort, 0)))  # Replace NAs with 0 to avoid breaks


ggplot(long_short_spread_brazil, aes(x = Date, y = CumulativeReturn)) +
  geom_line(color = "darkblue", size = 1.2) +
  labs(
    title = "Cumulative Return of Long–Short Momentum Strategy",
    x = "Date",
    y = "Cumulative Return"
  ) +
  theme_minimal()
```

```{r}
# Summary statistics for the long–short momentum strategy

# Remove NA values for performance metrics
spread_returns_brazil <- na.omit(long_short_spread_brazil$LongShort)

# Compute mean return, volatility, and Sharpe ratio
mean_return_brazil <- mean(spread_returns_brazil)
volatility_brazil <- sd(spread_returns_brazil)
sharpe_ratio_brazil <- mean_return_brazil / volatility_brazil

# Display metrics in a nicely formatted table
data.frame(
  Mean = mean_return_brazil,
  Volatility = volatility_brazil,
  Sharpe_Ratio = sharpe_ratio_brazil
) %>%
  knitr::kable(digits = 4, caption = "Performance Metrics for Long–Short Momentum Strategy")
```

### Step 11 – Regression Analysis

```{r}
# Clean data to avoid NA issues
regression_data_brazil <- final_data_brazil %>%
  filter(!is.na(ForwardReturn), !is.na(Decile))

# Regress forward returns on deciles
model_brazil <- lm(ForwardReturn ~ Decile, data = regression_data_brazil)

# View regression output
summary(model_brazil)
```

|     |     |     |     |     |
|-----|-----|-----|-----|-----|
|     |     |     |     |     |
|     |     |     |     |     |

## Country 2: Mexico 🇲🇽

### Step 1 – Load and Prepare Mexico Data

```{r}
# begin by importing historical price data for Mexican stocks 

mexico <- read_excel("Mexico_MEXBOL_Data.xlsx", na = c("N/A", "#N/A")) 
```

### Step 2 – Format Dates and Columns

```{r}
mexico <- mexico %>%                  
  rename(Date = 1) %>%                # Rename first column to 'Date'
  mutate(Date = as.Date(Date)) %>%    # Convert to Date format
  arrange(Date)                       # Sort data chronologically
```

### Step 3 – Convert Price Columns to Numeric

```{r}
# Convert all price columns to numeric (excluding Date)
mexico_prices <- mexico %>%
  mutate(across(-Date, ~ as.numeric(.)))
```

## Step 4 – Calculate Momentum (t-2 / t-12 - 1)

```{r}

# Calculate momentum: (Price at t-2 / Price at t-12) - 1
momentum_mexico <- mexico_prices %>%
  select(-Date) %>%                                          # Exclude Date column
  mutate(across(everything(), ~ lag(., 2) / lag(., 12) - 1)) # Momentum formula

# Reattach the Date column
momentum_mexico <- bind_cols(Date = mexico$Date, momentum_mexico)
```

### Step 5 – Rank into Deciles

```{r}
# Reshape to long format and assign momentum deciles per date
momentum_deciles_mexico <- momentum_mexico %>%
  pivot_longer(-Date, names_to = "Stock", values_to = "Momentum") %>%
  group_by(Date) %>%
  mutate(
    Decile = ntile(Momentum, 10)  # Rank into 10 groups
  ) %>%
  ungroup()
```

### Step 6 – Calculate 1-Month Forward Returns

```{r}
# 1. Remove Date column and calculate 1-month forward returns
returns_forward_mexico <- mexico %>%
  select(-Date) %>%
  mutate(across(everything(), ~ as.numeric(.))) %>%
  mutate(across(everything(), ~ lead(.) / . - 1))

# 2. Add back the Date column
returns_forward_mexico <- bind_cols(Date = mexico$Date, returns_forward_mexico)
```

### Step 7 – Merge Momentum & Returns

```{r}
# Reshape forward returns to long format (one row per stock-date)
returns_long_mexico <- returns_forward_mexico %>%
  pivot_longer(-Date, names_to = "Stock", values_to = "ForwardReturn")

# Merge forward returns with momentum decile information
final_data_mexico <- momentum_deciles_mexico %>%
  left_join(returns_long_mexico, by = c("Date", "Stock"))
```

### Step 8 – Summary Statistics and Visualisation

```{r}
# Calculate average 1-month forward return per momentum decile (across all dates)
decile_summary_mexico <- final_data_mexico %>%
  group_by(Decile) %>%
  summarise(
    AvgForwardReturn = mean(ForwardReturn, na.rm = TRUE), # Mean return by decile
    Count = n()                                           # Number of observations 
  )
```

```{r}
# Bar chart of average forward return by momentum decile
ggplot(decile_summary_mexico, aes(x = Decile, y = AvgForwardReturn)) +
  geom_col(fill = "steelblue") +
  labs(title = "Average 1-Month Forward Return by Momentum Decile",
       x = "Momentum Decile",
       y = "Avg Forward Return") +
  theme_minimal()
```

### Step 9 – Identify Top and Bottom Decile Stocks (Latest Date)

```{r}
# Get the latest date available in the data
latest_date_mexico <- max(momentum_deciles_mexico$Date, na.rm = TRUE)

# Extract top decile (highest momentum) stocks for that date
top_decile_stocks_mexico <- momentum_deciles_mexico %>%
  filter(Date == latest_date_mexico, Decile == 10) %>%
  select(Stock, Momentum)

# Extract bottom decile (lowest momentum) stocks for that date
bottom_decile_stocks_mexico <- momentum_deciles_mexico %>%
  filter(Date == latest_date_mexico, Decile == 1) %>%
  select(Stock, Momentum)

# Print the stocks in each group
cat("Top Decile (Winners) on", latest_date_mexico, ":\n")
print(top_decile_stocks_mexico)

cat("\nBottom Decile (Losers) on", latest_date_mexico, ":\n")
print(bottom_decile_stocks_mexico)

```

### Step 10 – Long–Short Strategy

```{r}
# Calculate long–short strategy returns: Decile 10 (long) minus Decile 1 (short)
long_short_spread_mexico <- final_data_mexico %>%
  filter(Decile %in% c(1, 10)) %>%    # Keep only top and bottom deciles
  group_by(Date, Decile) %>%
  summarise(Return = mean(ForwardReturn, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Decile, values_from = Return, names_prefix = "Decile") %>%
  mutate(LongShort = Decile10 - Decile1) %>%   # Strategy return
  arrange(Date)
```

```{r}
# Plot cumulative return of long–short strategy over time
long_short_spread_mexico <- long_short_spread_mexico %>%
  mutate(CumulativeReturn = cumsum(replace_na(LongShort, 0)))  # Replace NAs with 0 to avoid breaks


ggplot(long_short_spread_mexico, aes(x = Date, y = CumulativeReturn)) +
  geom_line(color = "darkblue", size = 1.2) +
  labs(
    title = "Cumulative Return of Long–Short Momentum Strategy",
    x = "Date",
    y = "Cumulative Return"
  ) +
  theme_minimal()
```

```{r}
# Summary statistics for the long–short momentum strategy

# Remove NA values for performance metrics
spread_returns_mexico <- na.omit(long_short_spread_mexico$LongShort)

# Compute mean return, volatility, and Sharpe ratio
mean_return_mexico <- mean(spread_returns_mexico)
volatility_mexico <- sd(spread_returns_mexico)
sharpe_ratio_mexico <- mean_return_mexico / volatility_mexico

# Display metrics in a nicely formatted table
data.frame(
  Mean = mean_return_mexico,
  Volatility = volatility_mexico,
  Sharpe_Ratio = sharpe_ratio_mexico
) %>%
  knitr::kable(digits = 4, caption = "Performance Metrics for Long–Short Momentum Strategy")
```

### Step 11 – Regression Analysis

```{r}
# Clean data to avoid NA issues
regression_data_mexico <- final_data_mexico %>%
  filter(!is.na(ForwardReturn), !is.na(Decile))

# Regress forward returns on deciles
model_mexico <- lm(ForwardReturn ~ Decile, data = regression_data_mexico)

# View regression output
summary(model_mexico)
```

## Country 3: India

### Step 1 – Load and Prepare India Data

```{r}
# begin by importing historical price data for Indian stocks 

india <- read_excel("India_Nifty_Data.xlsx", na = "N/A")
```

### Step 2 – Format Dates and Columns

```{r}
india <- india %>%                  
  rename(Date = 1) %>%                # Rename first column to 'Date'
  mutate(Date = as.Date(Date)) %>%    # Convert to Date format
  arrange(Date)                       # Sort data chronologically
```

### Step 3 – Convert Price Columns to Numeric

```{r}
# Convert all price columns to numeric (excluding Date)
india_prices <- india %>%
  mutate(across(-Date, ~ as.numeric(.)))
```

## Step 4 – Calculate Momentum (t-2 / t-12 - 1)

```{r}

# Calculate momentum: (Price at t-2 / Price at t-12) - 1
momentum_india <- india_prices %>%
  select(-Date) %>%                                          # Exclude Date column
  mutate(across(everything(), ~ lag(., 2) / lag(., 12) - 1)) # Momentum formula

# Reattach the Date column
momentum_india <- bind_cols(Date = india$Date, momentum_india)
```

### Step 5 – Rank into Deciles

```{r}
# Reshape to long format and assign momentum deciles per date
momentum_deciles_india <- momentum_india %>%
  pivot_longer(-Date, names_to = "Stock", values_to = "Momentum") %>%
  group_by(Date) %>%
  mutate(
    Decile = ntile(Momentum, 10)  # Rank into 10 groups
  ) %>%
  ungroup()
```

### Step 6 – Calculate 1-Month Forward Returns

```{r}
# 1. Remove Date column and calculate 1-month forward returns
returns_forward_india <- india %>%
  select(-Date) %>%
  mutate(across(everything(), ~ as.numeric(.))) %>%
  mutate(across(everything(), ~ lead(.) / . - 1))

# 2. Add back the Date column
returns_forward_india <- bind_cols(Date = india$Date, returns_forward_india)
```

### Step 7 – Merge Momentum & Returns

```{r}
# Reshape forward returns to long format (one row per stock-date)
returns_long_india <- returns_forward_india %>%
  pivot_longer(-Date, names_to = "Stock", values_to = "ForwardReturn")

# Merge forward returns with momentum decile information
final_data_india <- momentum_deciles_india %>%
  left_join(returns_long_india, by = c("Date", "Stock"))
```

### Step 8 – Summary Statistics and Visualisation

```{r}
# Calculate average 1-month forward return per momentum decile (across all dates)
decile_summary_india <- final_data_india %>%
  group_by(Decile) %>%
  summarise(
    AvgForwardReturn = mean(ForwardReturn, na.rm = TRUE), # Mean return by decile
    Count = n()                                           # Number of observations 
  )
```

```{r}
# Bar chart of average forward return by momentum decile
ggplot(decile_summary_india, aes(x = Decile, y = AvgForwardReturn)) +
  geom_col(fill = "steelblue") +
  labs(title = "Average 1-Month Forward Return by Momentum Decile",
       x = "Momentum Decile",
       y = "Avg Forward Return") +
  theme_minimal()
```

### Step 9 – Identify Top and Bottom Decile Stocks (Latest Date)

```{r}
# Get the latest date available in the data
latest_date_india <- max(momentum_deciles_india$Date, na.rm = TRUE)

# Extract top decile (highest momentum) stocks for that date
top_decile_stocks_india <- momentum_deciles_india %>%
  filter(Date == latest_date_india, Decile == 10) %>%
  select(Stock, Momentum)

# Extract bottom decile (lowest momentum) stocks for that date
bottom_decile_stocks_india <- momentum_deciles_india %>%
  filter(Date == latest_date_india, Decile == 1) %>%
  select(Stock, Momentum)

# Print the stocks in each group
cat("Top Decile (Winners) on", latest_date_india, ":\n")
print(top_decile_stocks_india)

cat("\nBottom Decile (Losers) on", latest_date_india, ":\n")
print(bottom_decile_stocks_india)

```

### Step 10 – Long–Short Strategy

```{r}
# Calculate long–short strategy returns: Decile 10 (long) minus Decile 1 (short)
long_short_spread_india <- final_data_india %>%
  filter(Decile %in% c(1, 10)) %>%    # Keep only top and bottom deciles
  group_by(Date, Decile) %>%
  summarise(Return = mean(ForwardReturn, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Decile, values_from = Return, names_prefix = "Decile") %>%
  mutate(LongShort = Decile10 - Decile1) %>%   # Strategy return
  arrange(Date)
```

```{r}
# Plot cumulative return of long–short strategy over time
long_short_spread_india <- long_short_spread_india %>%
  mutate(CumulativeReturn = cumsum(replace_na(LongShort, 0)))  # Replace NAs with 0 to avoid breaks


ggplot(long_short_spread_india, aes(x = Date, y = CumulativeReturn)) +
  geom_line(color = "darkblue", size = 1.2) +
  labs(
    title = "Cumulative Return of Long–Short Momentum Strategy",
    x = "Date",
    y = "Cumulative Return"
  ) +
  theme_minimal()
```

```{r}
# Summary statistics for the long–short momentum strategy

# Remove NA values for performance metrics
spread_returns_india <- na.omit(long_short_spread_india$LongShort)

# Compute mean return, volatility, and Sharpe ratio
mean_return_india <- mean(spread_returns_india)
volatility_india <- sd(spread_returns_india)
sharpe_ratio_india <- mean_return_india / volatility_india

# Display metrics in a nicely formatted table
data.frame(
  Mean = mean_return_india,
  Volatility = volatility_india,
  Sharpe_Ratio = sharpe_ratio_india
) %>%
  knitr::kable(digits = 4, caption = "Performance Metrics for Long–Short Momentum Strategy")
```

### Step 11 – Regression Analysis

```{r}
# Clean data to avoid NA issues
regression_data_india <- final_data_india %>%
  filter(!is.na(ForwardReturn), !is.na(Decile))

# Regress forward returns on deciles
model_india <- lm(ForwardReturn ~ Decile, data = regression_data_india)

# View regression output
summary(model_india)
```

## Cross-Country Comparative Analysis: Brazil, Mexico and India

```{r}
# Create performance metrics summary table
metrics_comparison <- data.frame(
  Country = c("Brazil", "Mexico", "India"),
  Mean = c(mean_return_brazil, mean_return_mexico, mean_return_india),
  Volatility = c(volatility_brazil, volatility_mexico, volatility_india),
  Sharpe_Ratio = c(sharpe_ratio_brazil, sharpe_ratio_mexico, sharpe_ratio_india)
)

# Display with nice formatting
metrics_comparison %>%
  arrange(desc(Sharpe_Ratio)) %>%
  knitr::kable(digits = 4, caption = "Performance Metrics Comparison: Long–Short Momentum Strategy")
```

```{r}
# Add country label to each dataset
long_short_spread_brazil$Country <- "Brazil"
long_short_spread_mexico$Country <- "Mexico"
long_short_spread_india$Country <- "India"

# Combine all 3
combined_cumulative <- bind_rows(
  long_short_spread_brazil %>% select(Date, CumulativeReturn, Country),
  long_short_spread_mexico %>% select(Date, CumulativeReturn, Country),
  long_short_spread_india %>% select(Date, CumulativeReturn, Country)
)

# Plot all together
ggplot(combined_cumulative, aes(x = Date, y = CumulativeReturn, color = Country)) +
  geom_line(size = 1.2) +
  labs(
    title = "📈 Cumulative Return Comparison: Long–Short Momentum Strategy",
    x = "Date",
    y = "Cumulative Return"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
# Identify best performer by each metric 
best_sharpe <- metrics_comparison$Country[which.max(metrics_comparison$Sharpe_Ratio)]
best_mean <- metrics_comparison$Country[which.max(metrics_comparison$Mean)]
lowest_vol <- metrics_comparison$Country[which.min(metrics_comparison$Volatility)]

cat("
### Summary of Findings

- Brazil had a Sharpe ratio of ", round(sharpe_ratio_brazil, 3), ", with a mean return of ", round(mean_return_brazil, 4), ".
- Mexico had a Sharpe ratio of ", round(sharpe_ratio_mexico, 3), ", with a mean return of ", round(mean_return_mexico, 4), ".
- India had a Sharpe ratio of ", round(sharpe_ratio_india, 3), ", with a mean return of ", round(mean_return_india, 4), ".

### Ranking Overview

- Best risk-adjusted performance (Sharpe ratio): ", best_sharpe, "
- Highest average return: ", best_mean, "
- Lowest volatility: ", lowest_vol, "

Overall, the results indicate that ", best_sharpe, " offered the most favourable risk-adjusted return profile among the three markets analysed during the sample period.
")
```

```{r}
# Add country labels
decile_summary_brazil$Country <- "Brazil"
decile_summary_mexico$Country <- "Mexico"
decile_summary_india$Country <- "India"

# Combine all
combined_deciles <- bind_rows(
  decile_summary_brazil,
  decile_summary_mexico,
  decile_summary_india
)

# Plot faceted comparison
ggplot(combined_deciles, aes(x = Decile, y = AvgForwardReturn)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~ Country) +
  labs(title = "Average 1-Month Forward Return by Momentum Decile",
       x = "Momentum Decile",
       y = "Average Forward Return") +
  theme_minimal()
```

```{r}
# Create individual country tables with renamed columns
brazil_tbl <- decile_summary_brazil %>% select(Decile, Brazil = AvgForwardReturn)
mexico_tbl <- decile_summary_mexico %>% select(Decile, Mexico = AvgForwardReturn)
india_tbl  <- decile_summary_india  %>% select(Decile, India  = AvgForwardReturn)

# Merge all country tables on Decile
momentum_table <- full_join(brazil_tbl, mexico_tbl, by = "Decile") %>%
  full_join(india_tbl, by = "Decile") %>%
  arrange(Decile) %>%
  filter(!is.na(Decile))  

# Round values for readability
momentum_table <- momentum_table %>%
  mutate(across(-Decile, ~ round(., 4)))

# Display the final table
knitr::kable(momentum_table, caption = "Average 1-Month Forward Return by Momentum Decile Across Countries")
```

## Read and prepare MSCI EM benchmark

```{r}
# MSCI EM Benchmark 

em_raw <- read_excel("Benchmark_IndexData.xlsx")  

# Try to standardise columns
em <- em_raw %>%
  rename(Date = 1) %>%
  mutate(Date = as.Date(Date)) %>%
  arrange(Date)

#  if there’s a single numeric column besides Date, treat it as Level; else look for 'Return'
num_cols <- names(em)[sapply(em, is.numeric)]
num_cols <- setdiff(num_cols, "Date")

if (length(num_cols) >= 1 && !"Return" %in% names(em)) {
  # Use the first numeric column as Level
  lvl_col <- num_cols[1]
  em <- em %>%
    rename(Level = all_of(lvl_col)) %>%
    mutate(EM_Forward = lead(Level) / Level - 1) %>%  
    select(Date, EM_Forward)
} else if ("Return" %in% names(em)) {
  # If a Return column already exists and is contemporaneous t->t+1, rename it
  em <- em %>%
    rename(EM_Forward = Return) %>%
    select(Date, EM_Forward)
} else {
  stop("MSCI EM file: couldn’t find a usable Level or Return column.")
}
```

# 2) Run EM-alpha, excess returns, and plots

```{r}
analyze_vs_em <- function(ls_df, decile_df, country_label) {
  # ls_df: data.frame with Date, LongShort
  # decile_df: data.frame with Date, Dec10 (avg forward return of decile 10 each month)

  # Merge EM
  ls <- ls_df %>%
    select(Date, LongShort) %>%
    left_join(em, by = "Date") %>%
    filter(!is.na(LongShort), !is.na(EM_Forward))

  dec10 <- decile_df %>%
    left_join(em, by = "Date") %>%
    filter(!is.na(Dec10), !is.na(EM_Forward))

  # Regressions (monthly)
  capm_ls   <- lm(LongShort ~ EM_Forward, data = ls)
  capm_dec10<- lm(Dec10 ~ EM_Forward, data = dec10)

  # Annualisation helpers (assume monthly data)
  ann <- function(x) {
    m <- mean(x, na.rm = TRUE)
    s <- sd(x, na.rm = TRUE)
    list(
      mean_ann = (1 + m)^12 - 1,
      vol_ann  = s * sqrt(12),
      sharpe_ann = ifelse(s == 0, NA, ( (1 + m)^12 - 1 ) / (s * sqrt(12)))
    )
  }

  # Excess & IR for long-only
  dec10_excess <- dec10$Dec10 - dec10$EM_Forward
  IR_dec10 <- mean(dec10_excess, na.rm = TRUE) / sd(dec10_excess, na.rm = TRUE)

  # Pack summaries
  out <- list(
    country = country_label,
    ls_alpha = broom::tidy(capm_ls),
    ls_glance = broom::glance(capm_ls),
    dec10_alpha = broom::tidy(capm_dec10),
    dec10_glance = broom::glance(capm_dec10),
    ls_ann = ann(ls$LongShort),
    dec10_ann = ann(dec10$Dec10),
    dec10_ir = IR_dec10,
    ls_excess_cum = ls %>% mutate(CumExcess = cumprod(1 + (LongShort - EM_Forward)) - 1),
    dec10_excess_cum = dec10 %>% mutate(CumExcess = cumprod(1 + (Dec10 - EM_Forward)) - 1)
  )
  out
}
```

# 3) Build Decile-10 (long-only) series per country

```{r}
# Brazil Decile 10 avg forward return each month
dec10_brazil <- final_data_brazil %>%
  filter(Decile == 10) %>%
  group_by(Date) %>%
  summarise(Dec10 = mean(ForwardReturn, na.rm = TRUE), .groups = "drop")

# Mexico
dec10_mexico <- final_data_mexico %>%
  filter(Decile == 10) %>%
  group_by(Date) %>%
  summarise(Dec10 = mean(ForwardReturn, na.rm = TRUE), .groups = "drop")

# India
dec10_india <- final_data_india %>%
  filter(Decile == 10) %>%
  group_by(Date) %>%
  summarise(Dec10 = mean(ForwardReturn, na.rm = TRUE), .groups = "drop")

```

# 4) Run analysis vs MSCI EM for each country

```{r}
res_br <- analyze_vs_em(
  ls_df = long_short_spread_brazil,
  decile_df = dec10_brazil,
  country_label = "Brazil"
)

res_mx <- analyze_vs_em(
  ls_df = long_short_spread_mexico,
  decile_df = dec10_mexico,
  country_label = "Mexico"
)

res_in <- analyze_vs_em(
  ls_df = long_short_spread_india,
  decile_df = dec10_india,
  country_label = "India"
)

```

# 5) Summaries and tables (alphas, betas, IR)

```{r}
# Long–Short CAPM vs EM (alpha, beta, t-stats)
ls_tbl <- bind_rows(
  res_br$ls_alpha %>% mutate(Country = "Brazil"),
  res_mx$ls_alpha %>% mutate(Country = "Mexico"),
  res_in$ls_alpha %>% mutate(Country = "India")
) %>%
  select(Country, term, estimate, std.error, statistic, p.value) %>%
  mutate(term = recode(term, `(Intercept)` = "Alpha", `EM_Forward` = "Beta (EM)")) %>%
  tidyr::pivot_wider(
    names_from = term,
    values_from = c(estimate, std.error, statistic, p.value)
  )

knitr::kable(ls_tbl, digits = 4, caption = "Long–Short vs MSCI EM: Alpha/Beta (Monthly)")
```

```{r}
# Long-only Decile 10 CAPM vs EM + Information Ratio (excess)
dec10_tbl <- bind_rows(
  res_br$dec10_alpha %>% mutate(Country = "Brazil"),
  res_mx$dec10_alpha %>% mutate(Country = "Mexico"),
  res_in$dec10_alpha %>% mutate(Country = "India")
) %>%
  select(Country, term, estimate, std.error, statistic, p.value) %>%
  mutate(term = recode(term, `(Intercept)` = "Alpha", `EM_Forward` = "Beta (EM)")) %>%
  tidyr::pivot_wider(
    names_from = term,
    values_from = c(estimate, std.error, statistic, p.value)
  ) %>%
  left_join(
    data.frame(
      Country = c("Brazil","Mexico","India"),
      Info_Ratio = c(res_br$dec10_ir, res_mx$dec10_ir, res_in$dec10_ir)
    ),
    by = "Country"
  )

knitr::kable(dec10_tbl, digits = 4, caption = "Decile 10 vs MSCI EM: Alpha/Beta (Monthly) and Information Ratio")
```

```{r}
# Annualised performance (for context)
ann_tbl <- tibble::tibble(
  Country = c("Brazil","Mexico","India"),
  LS_Mean_Ann = sapply(list(res_br$res, res_mx$res, res_in$res), function(x) NA) # placeholder to keep shape
) %>% select(-LS_Mean_Ann) %>%
  mutate(
    LS_Mean_Ann = c(res_br$ls_ann$mean_ann, res_mx$ls_ann$mean_ann, res_in$ls_ann$mean_ann),
    LS_Vol_Ann  = c(res_br$ls_ann$vol_ann,  res_mx$ls_ann$vol_ann,  res_in$ls_ann$vol_ann),
    LS_Sharpe   = c(res_br$ls_ann$sharpe_ann, res_mx$ls_ann$sharpe_ann, res_in$ls_ann$sharpe_ann),
    Dec10_Mean_Ann = c(res_br$dec10_ann$mean_ann, res_mx$dec10_ann$mean_ann, res_in$dec10_ann$mean_ann),
    Dec10_Vol_Ann  = c(res_br$dec10_ann$vol_ann,  res_mx$dec10_ann$vol_ann,  res_in$dec10_ann$vol_ann),
    Dec10_Sharpe   = c(res_br$dec10_ann$sharpe_ann, res_mx$dec10_ann$sharpe_ann, res_in$dec10_ann$sharpe_ann)
  )

knitr::kable(ann_tbl %>% mutate(across(-Country, round, 4)),
             caption = "Annualised Performance (Monthly data)")
```

# 6) Cumulative **excess** plots vs MSCI EM

```{r}
# Long–Short excess vs EM
ls_excess_all <- bind_rows(
  res_br$ls_excess_cum %>% mutate(Country = "Brazil"),
  res_mx$ls_excess_cum %>% mutate(Country = "Mexico"),
  res_in$ls_excess_cum %>% mutate(Country = "India")
)

ggplot(ls_excess_all, aes(x = Date, y = CumExcess, colour = Country)) +
  geom_line(size = 1.1) +
  labs(title = "Cumulative Excess Return of Long–Short vs MSCI EM",
       x = "Date", y = "Cumulative Excess Return") +
  theme_minimal() +
  theme(legend.position = "bottom")

```

```{r}
# Decile 10 (long-only) excess vs EM
dec10_excess_all <- bind_rows(
  res_br$dec10_excess_cum %>% mutate(Country = "Brazil"),
  res_mx$dec10_excess_cum %>% mutate(Country = "Mexico"),
  res_in$dec10_excess_cum %>% mutate(Country = "India")
)

ggplot(dec10_excess_all, aes(x = Date, y = CumExcess, colour = Country)) +
  geom_line(size = 1.1) +
  labs(title = "Cumulative Excess Return of Decile 10 vs MSCI EM",
       x = "Date", y = "Cumulative Excess Return") +
  theme_minimal() +
  theme(legend.position = "bottom")

```

## Machine Learning Part

```{r}
country_list <- list(
  Brazil = final_data_brazil,
  Mexico = final_data_mexico,
  India  = final_data_india
)

results <- list()

for (country in names(country_list)) {
  
  cat("\n===== Machine Learning for", country, "=====\n")
  
  ml_data <- country_list[[country]] %>%
    arrange(Stock, Date) %>%
    group_by(Stock) %>%
    mutate(
      Momentum_lag1 = lag(Momentum, 1),
      TopDecileThis = ifelse(Decile == 10, 1, 0)
    ) %>%
    ungroup() %>%
    filter(!is.na(Momentum_lag1), !is.na(TopDecileThis))
  
  # Train/test split (time-based)
  dates_sorted <- sort(unique(ml_data$Date))
  split_point <- dates_sorted[floor(length(dates_sorted) * 0.8)]
  
  train_df <- ml_data %>% filter(Date <= split_point)
  test_df  <- ml_data %>% filter(Date > split_point)
  
  # Random Forest classifier
  set.seed(123)
  rf_class <- randomForest(
    as.factor(TopDecileThis) ~ Momentum_lag1,
    data = train_df,
    ntree = 500
  )
  
  # Predict on test set
  test_df$PredProb <- predict(rf_class, newdata = test_df, type = "prob")[, 2]
  
  # Evaluate AUC & Hit Rate
  auc_score <- auc(test_df$TopDecileThis, test_df$PredProb)
  hit_rate <- mean(
    (test_df$PredProb >= 0.5 & test_df$TopDecileThis == 1) |
    (test_df$PredProb < 0.5 & test_df$TopDecileThis == 0)
  )
  
  cat("AUC:", round(auc_score, 4), "\n")
  cat("Directional Accuracy (Hit Rate):", round(hit_rate, 4), "\n")
  
  # Build long–short portfolio
  test_df <- test_df %>%
    group_by(Date) %>%
    mutate(
      RankProb = rank(-PredProb, ties.method = "first"),
      N = n(),
      Position = case_when(
        RankProb <= 0.1 * N ~ 1,   
        RankProb > 0.9 * N  ~ -1,  
        TRUE ~ 0
      )
    ) %>%
    ungroup()
  
  pred_ls <- test_df %>%
    group_by(Date) %>%
    summarise(
      LongShort = mean(ForwardReturn * Position, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(CumulativeReturn = cumprod(1 + replace_na(LongShort, 0)) - 1)
  
  results[[country]] <- list(
    AUC = auc_score,
    HitRate = hit_rate,
    Portfolio = pred_ls
  )
  
  # Plot for each country
  print(
    ggplot(pred_ls, aes(x = Date, y = CumulativeReturn)) +
      geom_line(color = "darkgreen", size = 1.1) +
      labs(
        title = paste("Predicted Long–Short Portfolio (", country, ")", sep = ""),
        x = "Date", y = "Cumulative Return"
      ) +
      theme_minimal()
  )
}
# Combine all countries' portfolio results into one dataframe
combined_results <- bind_rows(
  lapply(names(results), function(country) {
    results[[country]]$Portfolio %>% mutate(Country = country)
  })
)

# Plot the combined chart
ggplot(combined_results, aes(x = Date, y = CumulativeReturn, colour = Country)) +
  geom_line(size = 1.1) +
  labs(
    title = "Predicted Long–Short Portfolio: All Countries",
    x = "Date",
    y = "Cumulative Return"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.title = element_blank()
  )
```

```{r}
# ML Portfolio vs Baseline & MSCI EM (ALL COUNTRIES)

library(randomForest)
library(pROC)
library(broom)
library(dplyr)
library(tidyr)
library(ggplot2)
library(zoo)

# Expectation: 'em' exists with columns: Date, EM_Forward (monthly forward return)
stopifnot(all(c("Date","EM_Forward") %in% names(em)))

ann_stats <- function(x) {
  x <- na.omit(x); m <- mean(x); s <- sd(x)
  tibble(
    Mean_M = m,
    Vol_M  = s,
    Sharpe_M = ifelse(s == 0, NA, m/s),
    Mean_Ann = (1+m)^12 - 1,
    Vol_Ann  = s*sqrt(12),
    Sharpe_Ann = ifelse(s == 0, NA, ((1+m)^12 - 1)/(s*sqrt(12)))
  )
}

ml_vs_baseline <- function(final_data_country, country_label = "Country",
                           top_frac = 0.10, split_frac = 0.8) {

  # 1) Features (t-1) & target at t (leakage-free)
  df <- final_data_country %>%
    arrange(Stock, Date) %>%
    group_by(Stock) %>%
    mutate(
      Momentum_lag1 = lag(Momentum, 1),
      TopDecile_t   = ifelse(Decile == 10, 1, 0)
    ) %>%
    ungroup() %>%
    filter(!is.na(Momentum_lag1), !is.na(TopDecile_t), !is.na(ForwardReturn))

  # 2) Time split
  dates_sorted <- sort(unique(df$Date))
  split_point  <- dates_sorted[floor(length(dates_sorted) * split_frac)]
  train_df <- df %>% filter(Date <= split_point)
  test_df  <- df %>% filter(Date >  split_point)

  # 3) Train classifier
  set.seed(123)
  rf <- randomForest(as.factor(TopDecile_t) ~ Momentum_lag1,
                     data = train_df, ntree = 500, mtry = 1, importance = TRUE)

  # 4) Predict & metrics
  test_df$PredProb <- predict(rf, newdata = test_df, type = "prob")[,2]
  auc_val  <- as.numeric(pROC::auc(test_df$TopDecile_t, test_df$PredProb))
  hit_rate <- mean((test_df$PredProb >= 0.5 & test_df$TopDecile_t == 1) |
                   (test_df$PredProb <  0.5 & test_df$TopDecile_t == 0))

  # 5) Predicted ML long–short (equal-weight each month)
  test_df <- test_df %>%
    group_by(Date) %>%
    mutate(
      N = n(),
      RankProb = rank(-PredProb, ties.method = "first"),
      Position = case_when(
        RankProb <= top_frac * N ~  1,
        RankProb >  (1 - top_frac) * N ~ -1,
        TRUE ~ 0
      )
    ) %>% ungroup()

  pred_ls <- test_df %>%
    group_by(Date) %>%
    summarise(LS_ML = mean(ForwardReturn * Position, na.rm = TRUE), .groups = "drop")

  # 6) Baseline long–short (Decile10 - Decile1) on same test window
  base_ls <- final_data_country %>%
    group_by(Date, Decile) %>%
    summarise(R = mean(ForwardReturn, na.rm = TRUE), .groups = "drop") %>%
    filter(Decile %in% c(1,10)) %>%
    pivot_wider(names_from = Decile, values_from = R, names_prefix = "D") %>%
    mutate(LS_Base = D10 - D1) %>%
    select(Date, LS_Base) %>%
    filter(Date > split_point)

  # 7) Merge with MSCI EM and compute CAPM alphas
  cmp <- pred_ls %>%
    full_join(base_ls, by = "Date") %>%
    left_join(em, by = "Date") %>%
    arrange(Date) %>%
    filter(!is.na(EM_Forward))

  capm_ml   <- lm(LS_ML   ~ EM_Forward, data = cmp)
  capm_base <- lm(LS_Base ~ EM_Forward, data = cmp)

  # 8) Tables
  stats_tbl <- bind_rows(
    ann_stats(cmp$LS_ML)   %>% mutate(Series = "ML Long–Short"),
    ann_stats(cmp$LS_Base) %>% mutate(Series = "Baseline Long–Short"),
    ann_stats(cmp$EM_Forward) %>% mutate(Series = "MSCI EM")
  ) %>% relocate(Series)

  alpha_tbl <- bind_rows(
    broom::tidy(capm_ml)   %>% mutate(Series = "ML Long–Short"),
    broom::tidy(capm_base) %>% mutate(Series = "Baseline Long–Short")
  ) %>%
    mutate(term = recode(term, `(Intercept)` = "Alpha", `EM_Forward` = "Beta (EM)")) %>%
    select(Series, term, estimate, std.error, statistic, p.value) %>%
    pivot_wider(names_from = term,
                values_from = c(estimate, std.error, statistic, p.value))

  # 9) Plot per country (compounded cum returns)
  plot_df <- cmp %>%
    transmute(
      Date,
      `ML Long–Short`        = cumprod(1 + replace_na(LS_ML,   0)) - 1,
      `Baseline Long–Short`  = cumprod(1 + replace_na(LS_Base, 0)) - 1,
      `MSCI EM`              = cumprod(1 + EM_Forward) - 1
    ) %>%
    pivot_longer(-Date, names_to = "Series", values_to = "CumReturn")

  p <- ggplot(plot_df, aes(x = Date, y = CumReturn, colour = Series)) +
    geom_line(size = 1.1) +
    labs(title = paste0(country_label, ": Cumulative Returns (Test Period)"),
         x = "Date", y = "Cumulative Return") +
    theme_minimal() + theme(legend.position = "bottom")

  list(
    country   = country_label,
    auc       = auc_val,
    hit_rate  = hit_rate,
    stats     = stats_tbl,
    alpha_tbl = alpha_tbl,
    plot      = p,
    ml_series = plot_df %>% filter(Series == "ML Long–Short") %>%
                 transmute(Date, CumReturn, Country = country_label),
    feat_imp  = as.data.frame(importance(rf)) %>% tibble::rownames_to_column("Feature")
  )
}

# ---------- Run for all three countries ----------
res_br <- ml_vs_baseline(final_data_brazil, country_label = "Brazil")
res_mx <- ml_vs_baseline(final_data_mexico, country_label = "Mexico")
res_in <- ml_vs_baseline(final_data_india,  country_label = "India")

# Print key metrics and tables
cat("Brazil — AUC:", round(res_br$auc, 4), " | Hit Rate:", round(res_br$hit_rate, 4), "\n")
knitr::kable(res_br$stats %>% mutate(across(where(is.numeric), ~round(.,4))),
             caption = "Brazil: Monthly & Annualised Performance (Test Period)")
knitr::kable(res_br$alpha_tbl %>% mutate(across(where(is.numeric), ~round(.,4))),
             caption = "Brazil: Alpha/Beta vs MSCI EM (Monthly CAPM)")
print(res_br$plot)

cat("Mexico — AUC:", round(res_mx$auc, 4), " | Hit Rate:", round(res_mx$hit_rate, 4), "\n")
knitr::kable(res_mx$stats %>% mutate(across(where(is.numeric), ~round(.,4))),
             caption = "Mexico: Monthly & Annualised Performance (Test Period)")
knitr::kable(res_mx$alpha_tbl %>% mutate(across(where(is.numeric), ~round(.,4))),
             caption = "Mexico: Alpha/Beta vs MSCI EM (Monthly CAPM)")
print(res_mx$plot)

cat("India  — AUC:", round(res_in$auc, 4), " | Hit Rate:", round(res_in$hit_rate, 4), "\n")
knitr::kable(res_in$stats %>% mutate(across(where(is.numeric), ~round(.,4))),
             caption = "India: Monthly & Annualised Performance (Test Period)")
knitr::kable(res_in$alpha_tbl %>% mutate(across(where(is.numeric), ~round(.,4))),
             caption = "India: Alpha/Beta vs MSCI EM (Monthly CAPM)")
print(res_in$plot)

# ---------- One combined ML-only cumulative plot (all countries) ----------
combined_ml <- bind_rows(res_br$ml_series, res_mx$ml_series, res_in$ml_series) %>%
  mutate(Country = factor(Country, levels = c("Brazil","Mexico","India")))

ggplot(combined_ml, aes(x = Date, y = CumReturn, colour = Country)) +
  geom_line(size = 1.1) +
  labs(title = "ML Long–Short: Cumulative Returns by Country (Test Period)",
       x = "Date", y = "Cumulative Return") +
  theme_minimal() + theme(legend.position = "bottom")


```

```{r}
# Predicted vs Actual (Hexbin) — OOS Return Forecast 
# Simple regression model: predict ForwardReturn_t from Momentum_{t-1}
# Produces a hexbin-style Predicted vs Actual plot + MAE/RMSE/R^2 per country

library(randomForest)
library(dplyr)
library(ggplot2)
library(scales)

pred_vs_actual_plot <- function(final_data_country, country_label = "Country", split_frac = 0.8) {
  df <- final_data_country %>%
    arrange(Stock, Date) %>%
    group_by(Stock) %>%
    mutate(Momentum_lag1 = dplyr::lag(Momentum, 1)) %>%
    ungroup() %>%
    filter(!is.na(Momentum_lag1), !is.na(ForwardReturn))

  # Time-based split
  dates_sorted <- sort(unique(df$Date))
  split_point  <- dates_sorted[floor(length(dates_sorted) * split_frac)]
  train_df <- df %>% filter(Date <= split_point)
  test_df  <- df %>% filter(Date >  split_point)

  # Random Forest REGRESSION (forecast ForwardReturn)
  set.seed(123)
  rf_reg <- randomForest(ForwardReturn ~ Momentum_lag1,
                         data = train_df, ntree = 500, mtry = 1)

  # OOS predictions
  test_df$Pred <- predict(rf_reg, newdata = test_df)

  # Metrics
  mae  <- mean(abs(test_df$ForwardReturn - test_df$Pred), na.rm = TRUE)
  rmse <- sqrt(mean((test_df$ForwardReturn - test_df$Pred)^2, na.rm = TRUE))
  r2   <- cor(test_df$ForwardReturn, test_df$Pred, use = "complete.obs")^2

  # Hexbin-style plot (use 2D binning)
  p <- ggplot(test_df, aes(x = Pred, y = ForwardReturn)) +
    stat_bin2d(bins = 40) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    scale_fill_gradient(name = "Count", low = "#c6efff", high = "#ff3b30") +
    labs(
      title = paste0(country_label, " — Predicted vs Actual (OOS)"),
      subtitle = paste0("MAE: ", round(mae, 4),
                        " | RMSE: ", round(rmse, 4),
                        " | R²: ", round(r2, 4)),
      x = "Predicted Next-Month Return",
      y = "Actual Next-Month Return"
    ) +
    theme_minimal()

  list(plot = p, mae = mae, rmse = rmse, r2 = r2)
}

# ---- Run for all three countries and print plots ----
p_br <- pred_vs_actual_plot(final_data_brazil, "Brazil")
p_mx <- pred_vs_actual_plot(final_data_mexico, "Mexico")
p_in <- pred_vs_actual_plot(final_data_india,  "India")

print(p_br$plot)
print(p_mx$plot)
print(p_in$plot)

```

## highest-impact ML upgrades

### 1) Walk-forward (expanding window) validation

```{r}
# ---- Walk-forward expanding-window for one country (RF classifier) ----
wf_ml <- function(df, start_frac = 0.5, step = 1, top_frac = 0.10) {
  df <- df %>%
    arrange(Stock, Date) %>%
    group_by(Stock) %>% mutate(Momentum_lag1 = lag(Momentum, 1)) %>% ungroup() %>%
    filter(!is.na(Momentum_lag1), !is.na(ForwardReturn), !is.na(Decile)) %>%
    mutate(TopDecile_t = as.integer(Decile == 10))

  dates <- sort(unique(df$Date))
  start_i <- floor(length(dates) * start_frac)
  out <- list(pred=NULL, ls=NULL, auc=NULL, hit=NULL)

  for (i in seq(from = start_i, to = length(dates) - 1, by = step)) {
    train_end <- dates[i]
    test_date <- dates[i + 1]  # 1-month embargo by construction

    tr <- df %>% filter(Date <= train_end)
    te <- df %>% filter(Date == test_date)

    set.seed(123)
    rf <- randomForest::randomForest(
      as.factor(TopDecile_t) ~ Momentum_lag1,
      data = tr, ntree = 500, mtry = 1
    )

    te$PredProb <- predict(rf, newdata = te, type = "prob")[,2]

    # AUC / hit for this test month
    out$auc <- c(out$auc, pROC::auc(te$TopDecile_t, te$PredProb))
    out$hit <- c(out$hit, mean((te$PredProb >= 0.5 & te$TopDecile_t == 1) |
                               (te$PredProb < 0.5  & te$TopDecile_t == 0)))

    # Long-short for this month (exact neutrality)
    te <- te %>% mutate(rk = rank(-PredProb, ties.method="first"),
                        N = n(),
                        pos = dplyr::case_when(
                          rk <= top_frac*N ~  1,
                          rk >  (1-top_frac)*N ~ -1,
                          TRUE ~ 0))

    ret_month <- mean(te$ForwardReturn[te$pos== 1], na.rm=TRUE) -
                 mean(te$ForwardReturn[te$pos==-1], na.rm=TRUE)

    out$ls <- dplyr::bind_rows(out$ls, tibble::tibble(Date=test_date, LS_ML=ret_month))
    out$pred <- dplyr::bind_rows(out$pred, te %>% dplyr::select(Date, Stock, PredProb))
  }
  out
}

```

Run per country:

```{r}
wf_br <- wf_ml(final_data_brazil)
wf_mx <- wf_ml(final_data_mexico)
wf_in <- wf_ml(final_data_india)

```

## 2) Transaction costs & turnover (make it realistic)

Add net performance after costs. Example for a long–short series `ls_df` and **per-side cost** `tc` (e.g., 25 bps = 0.0025):

```{r}
# ls_positions is the per-name Position matrix already build each month

turnover_from_positions <- function(pos_df) {
  # pos_df: Date, Stock, Position in {-1,0,1}
  pos_df %>%
    arrange(Stock, Date) %>%
    group_by(Stock) %>% mutate(prev = dplyr::lag(Position, 1)) %>% ungroup() %>%
    mutate(chg = abs((Position - dplyr::coalesce(prev, 0)))) %>%
    group_by(Date) %>% summarise(Turnover = mean(chg, na.rm=TRUE)/2, .groups="drop")
    # /2 since a round-trip (0->1->0) counts two one-way trades
}

# Join with returns and subtract costs
apply_costs <- function(ls_ret_df, turnover_df, tc = 0.0025) {
  ls_ret_df %>%
    left_join(turnover_df, by="Date") %>%
    mutate(Turnover = dplyr::coalesce(Turnover, 0),
           Net = LS_ML - tc * Turnover) %>%
    mutate(Cum_Gross = cumprod(1 + dplyr::coalesce(LS_ML,0)) - 1,
           Cum_Net   = cumprod(1 + dplyr::coalesce(Net,0))   - 1)
}

```

## 3) Tune the portfolio **breadth** (top/bottom %) out-of-sample

```{r}
breadth_test <- function(test_df) {
  fracs <- c(0.05, 0.10, 0.20)
  out <- purrr::map_dfr(fracs, function(f){
    tmp <- test_df %>%
      group_by(Date) %>%
      mutate(N=n(), r=rank(-PredProb, ties.method="first"),
             pos = case_when(r <= f*N ~ 1,
                             r > (1-f)*N ~ -1, TRUE ~ 0)) %>%
      ungroup() %>%
      group_by(Date) %>%
      summarise(LS = mean(ForwardReturn[pos==1], na.rm=TRUE) -
                     mean(ForwardReturn[pos==-1], na.rm=TRUE), .groups="drop")
    tibble::tibble(Frac=f,
                   Mean=mean(tmp$LS, na.rm=TRUE),
                   Vol=sd(tmp$LS, na.rm=TRUE),
                   Sharpe=Mean/Vol)
  })
  out
}

```

## 4) Ensemble = ML score + classic momentum rank

```{r}
# Within each date, z-score both signals then average
ensemble_scores <- function(df_with_preds) {
  df_with_preds %>%
    group_by(Date) %>%
    mutate(
      z_ml  = scale(PredProb),
      z_mom = scale(Momentum_lag1),
      Score = 0.5 * as.numeric(z_ml) + 0.5 * as.numeric(z_mom)
    ) %>%
    ungroup()
}

```

## 5) Statistical significance of alphas (Newey–West)

```{r}
# install.packages(c("sandwich","lmtest"))
library(sandwich); library(lmtest)
nw_alpha <- function(ret, mkt){
  m <- lm(ret ~ mkt)
  co <- coeftest(m, vcov = NeweyWest(m, lag = 3, prewhite = FALSE))
  co  # alpha is the intercept row
}
# Example: nw_alpha(cmp$LS_ML, cmp$EM_Forward)

```

## 6) Class imbalance & threshold tuning

```{r}
# --- Balanced RF diagnostics (robust) ---
library(pROC)

# 1) Pick the target (0/1) and probability columns that actually exist
target_col <- if ("TopDecile_t" %in% names(test_df)) {
  "TopDecile_t"
} else if ("TopDecileThis" %in% names(test_df)) {
  "TopDecileThis"
} else {
  stop("Could not find TopDecile_t / TopDecileThis in test_df.")
}

prob_col <- if ("PredProb_bal" %in% names(test_df)) {
  "PredProb_bal"
} else if ("PredProb" %in% names(test_df)) {
  "PredProb"
} else {
  stop("Could not find PredProb_bal / PredProb in test_df.")
}

# 2) Keep only valid rows and coerce the label to a 0/1 factor
diag_df <- test_df |>
  dplyr::select(Label = dplyr::all_of(target_col),
                Score = dplyr::all_of(prob_col)) |>
  dplyr::filter(!is.na(Label), !is.na(Score))

# Quick sanity check 
# print(table(diag_df$Label))

if (nrow(diag_df) == 0) stop("No valid rows for ROC after NA filtering.")

# Ensure factor with levels "0","1" (controls < cases)
diag_df$Label <- factor(as.character(diag_df$Label), levels = c("0","1"))

# 3) ROC with Youden-optimal threshold
roc_obj <- pROC::roc(
  response  = diag_df$Label,
  predictor = diag_df$Score,
  levels    = c("0","1"),
  direction = "<",
  quiet     = TRUE
)

best <- pROC::coords(roc_obj, x = "best", best.method = "youden",
                     ret = c("threshold","sensitivity","specificity"))
best_thr <- as.numeric(best["threshold"])

# 4) Metrics at tuned threshold
pred_label <- as.integer(diag_df$Score >= best_thr)  # 1 if score >= thr else 0
auc_bal    <- as.numeric(pROC::auc(roc_obj))
hit_bal    <- mean((pred_label == 1 & diag_df$Label == "1") |
                   (pred_label == 0 & diag_df$Label == "0"))

cat("Balanced RF — AUC:", round(auc_bal,4),
    "| Hit Rate (tuned):", round(hit_bal,4),
    "| Threshold:", round(best_thr,3), "\n")

# Optional: confusion matrix + precision/recall/F1
cm <- table(Actual = diag_df$Label,
            Predicted = factor(pred_label, levels = c(0,1), labels = c("0","1")))
precision <- ifelse(sum(cm[ , "1"]) == 0, NA, cm["1","1"] / sum(cm[ , "1"]))
recall    <- cm["1","1"] / sum(cm["1", ])
f1        <- ifelse(is.na(precision) | is.na(recall) | (precision+recall)==0,
                    NA, 2*precision*recall/(precision+recall))
cat("Precision:", round(precision,4),
    "| Recall:", round(recall,4),
    "| F1:", round(f1,4), "\n")

```

```{r}
# Walk-Forward Summary + Costs + Breadth (ALL COUNTRIES) 

# 1) Walk-forward expanding window, returns predictions, positions, AUC/Hit
wf_ml_full <- function(df, start_frac = 0.5, step = 1, top_frac = 0.10) {
  panel <- df %>%
    arrange(Stock, Date) %>%
    group_by(Stock) %>% mutate(Momentum_lag1 = dplyr::lag(Momentum, 1)) %>% ungroup() %>%
    filter(!is.na(Momentum_lag1), !is.na(ForwardReturn), !is.na(Decile)) %>%
    mutate(TopDecile = as.integer(Decile == 10))

  dates <- sort(unique(panel$Date))
  start_i <- floor(length(dates) * start_frac)

  out <- list(ls = NULL, auc_month = NULL, positions = NULL, pred = NULL)

  for (i in seq(from = start_i, to = length(dates) - 1, by = step)) {
    train_end <- dates[i]
    test_date <- dates[i + 1]

    tr <- panel %>% filter(Date <= train_end)
    te <- panel %>% filter(Date == test_date)

    set.seed(123)
    rf <- randomForest::randomForest(
      as.factor(TopDecile) ~ Momentum_lag1,
      data = tr, ntree = 500, mtry = 1
    )

    te$PredProb <- predict(rf, newdata = te, type = "prob")[,2]
    auc_val <- as.numeric(pROC::auc(te$TopDecile, te$PredProb))
    hit_val <- mean((te$PredProb >= 0.5 & te$TopDecile == 1) |
                    (te$PredProb <  0.5 & te$TopDecile == 0))

    te <- te %>%
      mutate(
        N  = n(),
        rk = rank(-PredProb, ties.method = "first"),
        Position = case_when(
          rk <= top_frac * N ~  1,
          rk >  (1 - top_frac) * N ~ -1,
          TRUE ~ 0
        )
      )

    ret_month <- mean(te$ForwardReturn[te$Position ==  1], na.rm = TRUE) -
                 mean(te$ForwardReturn[te$Position == -1], na.rm = TRUE)

    out$ls        <- bind_rows(out$ls, tibble(Date = test_date, LS_Gross = ret_month))
    out$auc_month <- bind_rows(out$auc_month, tibble(Date = test_date, AUC = auc_val, Hit = hit_val))
    out$positions <- bind_rows(out$positions, te %>% select(Date, Stock, Position))
    out$pred      <- bind_rows(out$pred, te %>% select(Date, Stock, PredProb, ForwardReturn))
  }
  out
}

# 2) Turnover from positions & apply costs to get net series
turnover_from_positions <- function(pos_df) {
  pos_df %>%
    arrange(Stock, Date) %>%
    group_by(Stock) %>% mutate(prev = dplyr::lag(Position, 1)) %>% ungroup() %>%
    mutate(chg = abs(Position - dplyr::coalesce(prev, 0))) %>%
    group_by(Date) %>%
    summarise(Turnover = mean(chg, na.rm = TRUE) / 2, .groups = "drop")
}

compute_net_perf <- function(wf_obj, tc = 0.0025) {  # tc = 25 bps per one-way
  turn <- turnover_from_positions(wf_obj$positions)
  perf <- wf_obj$ls %>%
    left_join(turn, by = "Date") %>%
    mutate(Turnover = dplyr::coalesce(Turnover, 0),
           LS_Net   = LS_Gross - tc * Turnover,
           Cum_Gross = cumprod(1 + dplyr::coalesce(LS_Gross, 0)) - 1,
           Cum_Net   = cumprod(1 + dplyr::coalesce(LS_Net,   0)) - 1)
  list(
    perf   = perf,
    gross  = ann_stats(wf_obj$ls$LS_Gross),
    net    = ann_stats(perf$LS_Net)
  )
}

# 3) Breadth robustness using stored predictions
breadth_from_preds <- function(pred_panel, fracs = c(0.05, 0.10, 0.20)) {
  purrr::map_dfr(fracs, function(f){
    tmp <- pred_panel %>%
      group_by(Date) %>%
      mutate(N = n(),
             rk = rank(-PredProb, ties.method = "first"),
             pos = case_when(rk <= f*N ~  1,
                             rk > (1-f)*N ~ -1,
                             TRUE ~ 0)) %>%
      ungroup() %>%
      group_by(Date) %>%
      summarise(LS = mean(ForwardReturn[pos==1], na.rm=TRUE) -
                     mean(ForwardReturn[pos==-1], na.rm=TRUE),
                .groups = "drop")
    tibble(
      Frac = f,
      Mean_M = mean(tmp$LS, na.rm=TRUE),
      Vol_M  = sd(tmp$LS, na.rm=TRUE),
      Sharpe_M = ifelse(sd(tmp$LS, na.rm=TRUE)==0, NA, mean(tmp$LS, na.rm=TRUE)/sd(tmp$LS, na.rm=TRUE)),
      Mean_Ann = (1+Mean_M)^12 - 1,
      Vol_Ann  = Vol_M * sqrt(12),
      Sharpe_Ann = ifelse(Vol_M == 0, NA, ((1+Mean_M)^12 - 1)/(Vol_M*sqrt(12)))
    )
  })
}

# Run for all three countries 
wf_br <- wf_ml_full(final_data_brazil, start_frac = 0.5, top_frac = 0.10)
wf_mx <- wf_ml_full(final_data_mexico, start_frac = 0.5, top_frac = 0.10)
wf_in <- wf_ml_full(final_data_india,  start_frac = 0.5, top_frac = 0.10)

# Step 1: Average AUC/Hit across test months
auc_hit_tbl <- tibble::tibble(
  Country = c("Brazil","Mexico","India"),
  AUC_Avg = c(mean(wf_br$auc_month$AUC, na.rm=TRUE),
              mean(wf_mx$auc_month$AUC, na.rm=TRUE),
              mean(wf_in$auc_month$AUC, na.rm=TRUE)),
  Hit_Avg = c(mean(wf_br$auc_month$Hit, na.rm=TRUE),
              mean(wf_mx$auc_month$Hit, na.rm=TRUE),
              mean(wf_in$auc_month$Hit, na.rm=TRUE))
)
knitr::kable(auc_hit_tbl %>% mutate(across(-Country, ~round(.,4))),
             caption = "Walk-Forward: Average AUC and Hit Rate (Test Months)")

#Step 2: Net vs Gross Sharpe (with costs) + Plot cum returns
tc <- 0.0025  # 25 bps per one-way trade; adjust as needed

net_br <- compute_net_perf(wf_br, tc = tc)
net_mx <- compute_net_perf(wf_mx, tc = tc)
net_in <- compute_net_perf(wf_in, tc = tc)

gn_tbl <- bind_rows(
  net_br$gross %>% mutate(Country="Brazil", Series="Gross"),
  net_br$net   %>% mutate(Country="Brazil", Series="Net"),
  net_mx$gross %>% mutate(Country="Mexico", Series="Gross"),
  net_mx$net   %>% mutate(Country="Mexico", Series="Net"),
  net_in$gross %>% mutate(Country="India",  Series="Gross"),
  net_in$net   %>% mutate(Country="India",  Series="Net")
) %>% relocate(Country, Series)

knitr::kable(gn_tbl %>% mutate(across(where(is.numeric), ~round(.,4))),
             caption = paste0("Walk-Forward Long–Short: Gross vs Net (tc = ", tc, " per side)"))

# Cumulative Gross vs Net plots (one per country)
plot_gn <- function(perf_df, title_prefix){
  ggplot(perf_df %>% select(Date, Cum_Gross, Cum_Net) %>%
           pivot_longer(-Date, names_to="Series", values_to="Cum"),
         aes(x=Date, y=Cum, colour=Series)) +
    geom_line(size=1.1) +
    labs(title = paste0(title_prefix, " — Walk-Forward Cumulative Returns"),
         x="Date", y="Cumulative Return") +
    theme_minimal() + theme(legend.position="bottom")
}
print(plot_gn(net_br$perf, "Brazil"))
print(plot_gn(net_mx$perf, "Mexico"))
print(plot_gn(net_in$perf, "India"))

# Step 3: Breadth robustness (5%, 10%, 20%)
breadth_tbl <- bind_rows(
  breadth_from_preds(wf_br$pred) %>% mutate(Country="Brazil"),
  breadth_from_preds(wf_mx$pred) %>% mutate(Country="Mexico"),
  breadth_from_preds(wf_in$pred)  %>% mutate(Country="India")
) %>% relocate(Country, Frac)

knitr::kable(breadth_tbl %>% mutate(across(where(is.numeric), ~round(.,4))),
             caption = "Breadth Robustness (Top/Bottom Fraction): Monthly & Annualised Stats")

```
